---
title: "Pearson's r Correlation Template"
author: Jacob Coutts
format: docx
editor: visual
---

-   This is the template you will use to run a Pearson's r correlation analysis.
-   In this template, the predictor variable is labeled 'predictor' and the outcome variable is labeled 'outcome'.
-   You will be using the template to run a correlation analysis with your data.

## Step 1: Load required packages.

**You do not need to change anything in the chunk of code below. Simply run the code to load the libraries.**

```{r}
library(psych)
library(ggplot2)
library(dplyr)
```

## Step 2: Reading in your data.

**Remember to change this to the path/name for YOUR data!**

```{r}
DF <- read.table("correlation_file_name.csv", header= TRUE, sep=",")
```

## Step 3: Exploring your data.

**View summary statistics of your dataset.**

```{r}
psych::describeBy(DF)
```

## Step 4: Examine statistical assumptions (diagnostics).

We will make a quick scatter plot to assess homoscedasticity and linearity. Look for funnel shapes and curves. Remember: Below, change predictorname and outcomename to the name of the variables in YOUR dataset. If you don't use them exactly, you will generate an error (because R will be looking for objects that don't exist).

```{r}
ggplot(DF, aes(x=predictorname, y=outcomename)) + 
  geom_point() + 
  geom_jitter() + 
  ggtitle("title") + 
  labs(y="outcome name", x = "predictor name")
```

## Step 5: Run the analysis and make an inference.

**Remember, DF must be the name of your dataset in R and predictor/outcomename must be the names of those variables.**

```{r}
cor.test(DF$predictorname,DF$outcomename) 
```

## Step 6: Visualize the results.

**Remember to change the following information IN THE CODE CHUNK BELOW.** These are the descriptions for what you will need to fill in where there is green text (character strings) in the code block below.

1.  Your graph title in the ggtitle("") command: This can be anything you want within reason. Make sure it stays within the ("").
2.  Your y-axis title in the labs(y = "") command): This can be anything you want. For example, if your variable name is TimeSleep you can name the axis "Time Spent Sleeping". Make sure it stays within the ("").
3.  Your x-axis title in the labs(x = "") command): This can be anything you want. For example, if your variable name is ExamScore you can name the axis "Score on Exam". Make sure it stays within the ("").

```{r}
ggplot(DF, aes(x=predictorname, y=outcomename)) + 
  geom_point() + 
  geom_jitter() + 
  ggtitle("title") + 
  labs(y="outcome", x = "predictor")
```

## Step 7: Render the file.

You will see a blue arrow with the word 'Render' next to it at the top of the screen. Click this icon.

### End of script
